\vspace{1.0cm} \Large \bf Integrating influenza antigenic dynamics with molecular evolution

### 2013-07-23 response.tex

The comparison in Table 1 is crucial, but the model descriptions are very hard to understand until after reading the Methods carefully. The difficulty is because the models are described in technical terms (dimensionality, location prior, serum effects, virus effects). The most confusing part is the location prior. My understanding (and if this is wrong then things really need to be clarified) is that uninformed refers to making the predictions just from HI data, drift refers to making the predictions from HI data + date of isolation, and diffusion / drift refers to making the predictions from HI data + date of isolation + phylogenetic relationship. If the models were named in terms of the data used to make the predictions rather than in terms of the prior, it would be easier to understand intuitively. For instance, a reader could then simply see, ``OK, they included the isolation date in the model and the predictions improved by this much.'' I recognize that the prior plays a role in addition to the data, but I feel that a more intuitive naming scheme should be considered with detailed descriptions in the Methods. There also might be a more intuitive way to denote serum effects and virus effects, although this is less confusing than the location prior.

The first figure of the paper (Figure 1) is basically methodological rather than presenting a biological result. Although the data in this figure may be important, making it the first thing the reader encounters makes the paper less accessible to readers primarily interested in the biological results.

\textbf{We agree and have revised the manuscript to remove this figure.  Instead of the figure we report Pearson's correlations between antigenic distance and error ($r = 0.098$) and between year and error ($r = -0.007$).}

@@

I found the section fitting diffusion coefficients (in different branches) to the antigenic maps confusing -- if only because strict diffusion (in one or two dimensions) has no advection term and therefore no tendency to move in one direction or another; whereas the data clearly show an advective tendency (moving to the ``right'' over time in antigenic dimension 1). I presume the authors attempt to deal with this by imposing a prior (eq 12) for the location of a virus based on its location in a molecular phylogeny. But this would seem to conflate antigenic drift, which the authors wish to quantify using a diffusion coefficient, with genetic mutations (many of which are driven by selection for antigenic escape). In other words, this hybrid approach of assuming a prior based on a strain's genotype, and then fitting a diffusion coefficient to summarize strength of drift, does not seem to to be a pure measure of drift. What exactly the fitted diffusion coefficient means is extremely unclear to me, in any formal sense, as the procedure is so ad-hoc.

How do the antigenic maps shown compare to BMDS applied to pairwise aa hamming distances alone? How many serious discrepancies that are outside the confidence intervals on map reconstruction? In other words, the authors should demonstrate that the HI data actually provide substantively different results for antigenic grouping than genetic data do alone. This was an important contention of Smith et al (2004), but it was based on the position of only a handful of viruses in that paper. Do the authors have more convincing evidence in this study that the antigenic data are providing substantial extra information beyond the genetic data?

\textbf{We have revised the manuscript to include a thorough analysis of the relationship between genetic distance and antigenic distance.  This is new figure 1.  Rather than constructing BMDS maps with pairwise AA distances, we take a more direct approach and compare pairwise antigenic distance from a BMDS model using only HI data (model 2 in Table 1) and pairwise genetic distances.  We examine both pairwise AA distance and pairwise phylogenetic distance, as phylogenetic distance was the basis for our evolutionary diffusion model.  We find fairly modest correlations in most cases case (between 0.10 for B/Yam and 0.68 for A/H3N2).  Genetic data provides some predictive power for antigenic distance, but is a rather weak predictor alone.  This was a big part of our rationale for attempting a joint genetic / antigenic model.}

### 2013-07-23 flux.tex
Fix #25 - antigenic vs genetic distances. Include new figure 1.

Previous work on influenza antigenic and genetic evolution has examined antigenic and genetic patterns in isolation, and then compared the results to assess congruence of these two aspects of evolution \cite{Hay01, Smith04, Russell08}. 
Here, we simultaneously model antigenic and genetic evolution by adopting an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of the phylogenetic tree according to a Brownian motion process (see Materials and methods).
The phylogenetic diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
The antigenic diffusion process includes both systematic drift with time and covariance induced by phylogenetic proximity.
We examine the effects of including only systematic drift (model 6; Table \ref{errortable}) and systematic drift plus phylogenetic diffusion (model 7; Table \ref{errortable}), finding a small increase in predictive accuracy of between 0.02 and 0.08 log$_2$ HI titers when both processes are included.
The systematic drift process informs virus and serum locations by dates of isolation and the phylogenetic diffusion process informs virus locations by genetic sequences.
Thus, in these models, antigenic locations are inferred using both genetic data and HI data and will differ from locations inferred from HI data alone.
If HI data is rich, then we expect minor differences in antigenic locations with the inclusion of genetic data (as may be the case for A/H3N2), while if HI data is spare, then we expect genetic data to play a larger role in determining antigenic locations (as may be the case for B/Vic and B/Yam).

>>>>

Previous work on influenza antigenic and genetic evolution has shown that antigenic distance accumulates with increasing genetic distance \cite{Hay01, Smith04, Russell08}.
Here, we examine pairwise relationships between viruses and observe a correlation between amino acid mutations and antigenic distance (Figure~\ref{seq_grid}) and a similar correlation between phylogenetic distance and antigenic distance (Figure~\ref{seq_grid}).
Thus, genetic relationships between viruses provide some predictive power to estimate antigenic distances in the absence of HI data.
However, the magnitudes of the coefficients of determination $R^2$ are low (Figure~\ref{seq_grid}), suggesting that genetic relationships alone will not completely resolve antigenic distances.

%%% map %%%
\begin{figure}[h]
	\centering		
	\includegraphics[width=1.0\textwidth]{figures/seq_grid}
	\caption{\textbf{Pairwise correlations between genetic distance, measured as amino acid mutations or as phylogenetic distance, and antigenic distance for influenza A/H3N2, A/H1N1, B/Vic and B/Yam.} 
	The top row shows correlations between number of amino acid mutations and average antigenic distance between 10,000 random pairs of viruses.
	The bottom row shows correlations between average phylogenetic distance, measured in terms of years, and average antigenic distance between 10,000 random pairs of viruses.
	Dashed lines show linear model fits, with $R^2$ and slope noted, while solid lines show LOESS fits.
	Antigenic distances derive from model 2 of Table \ref{errortable}.
	} 
	\label{seq_grid} 
\end{figure}

Consequently, we seek to flexibly incorporate genetic data by simultaneously modeling antigenic and genetic evolution through an evolutionary diffusion process \cite{Lemey10}, wherein a virus's antigenic character state evolves along branches of the phylogenetic tree according to a Brownian motion process (see Materials and methods).
The phylogenetic diffusion process acts as a prior on virus locations, so that genetically similar viruses are expected to share similar antigenic locations.
The antigenic diffusion process includes both systematic drift with time and covariance induced by phylogenetic proximity.
We examine the effects of including only systematic drift (model 6; Table \ref{errortable}) and systematic drift plus phylogenetic diffusion (model 7; Table \ref{errortable}), finding a small increase in predictive accuracy of between 0.02 and 0.08 log$_2$ HI titers when both processes are included.
The systematic drift process informs virus and serum locations by dates of isolation and the phylogenetic diffusion process informs virus locations by genetic sequences.
Thus, in these models, antigenic locations are inferred using both genetic data and HI data and will differ from locations inferred from HI data alone.
If HI data is rich, then we expect minor differences in antigenic locations with the inclusion of genetic data (as may be the case for A/H3N2), while if HI data is spare, then we expect genetic data to play a larger role in determining antigenic locations (as may be the case for B/Vic and B/Yam).
